---
title: "Statistics with R - Assignment1"
author: "Ville Lehtonen"
output: html_document
date: "24.4.2020"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(knitr)
library(readxl)
library(tidyverse)

# If the data files are in a separate folder, uncomment the below code:
#current_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
#data_file_path <- paste(current_dir, '/data', sep = "")
#knitr::opts_knit$set(root.dir = data_file_path)
```

## Question 1:
##### What was good in the data:
 - no cell formatting was used (i.e. colors)  
 
 - comments were put in separate columns  
 
 - variable names were clear and camelCase was used mostly (although in some columns this was not the case)  
 
 - Data was split into different tables and samples and counts tables were connected by the sample value  
 
##### What could be improved:
 - The counts data could have been in a 'long' format i.e. there should have been a column called 'sample' and 'count', instead of having multiple samples as separate columns  
 
 - The 'bad data' (with i.e. missing values) could have been labeled in a new column that would have had just TRUE / FALSE values indicating whether this is good or bad data. This would ease the filtering of data  

 - in 'samples' table, the time columns could have been split into hours and minutes columns so that they would only include numerical values  
 
- in 'samples' table, there was a column called 'duplicate' that contained french (Oui / Non). Instead it would be simpler if it would contain only a logical TRUE / FALSE value  

- Metadata was missing  


## Question 2:
Create objects for counts, samples, and communities data
```{r question 2}
counts <- read_xlsx('lait_caille_data.xlsx',
                    sheet = 'counts',
                    col_names = TRUE)

samples <- read_xlsx('lait_caille_data.xlsx',
                    sheet = 'samples',
                    col_names = TRUE)

communities <- read_xlsx('lait_caille_data.xlsx',
                    sheet = 'communities',
                    col_names = TRUE)

```


## Question 3:
Split counts data and tidy it
```{r question 3}
# Split counts data into taxonomy and counts objects (both contain rankID)
taxonomy <- select(counts,taxlevel, rankID, taxon)
counts <- select(counts, -taxlevel, -taxon)

# Transform counts object into a tidy format with 3 columns (rankID, sample, count)
counts_tidy <- gather(data = counts,
                      key = sample,
                      value = count,
                      -rankID)
```


## Question 4:
Perform checks
```{r question 4}
# Gather all unique sample IDs from 'counts' & 'samples'
counts_IDs <- unique(unlist(counts_tidy$sample, " "))
samples_IDs <- samples$sample

# Check that both IDs vectors contain the same objects
print(c("Check that all IDs of counts are in samples:", all(counts_IDs %in% samples_IDs)))
print(c("Check that all IDs of samples are in counts:", all(samples_IDs %in% counts_IDs)))

# Gather unique community names from 'communities' and 'samples'
original_names <- communities$original
samples_communities <- unique(unlist(samples$community, " "))

# Check if all original names are included in communities in 'samples' and vice versa 
print(c("Check that all original names are in the communities of samples:", all(original_names %in% samples_communities)))
print(c("Check that all communities of samples are in the original names:", all(samples_communities %in% original_names))) #-> FALSE

# Gather the missing communities (= surplus or have missing values) into a list
missing_communities <- samples_communities[!samples_communities %in% original_names]
```

Comments: It seems that not all communities found in samples are present in the original names



## Question 5:
Filter rows where community is in senegal and taxonomy level is 6
```{r question 5}
# Add community to 'counts' by sample ID from 'samples'
counts_tidy_merged <- merge(counts_tidy, samples[, c("sample", "community")], 
                     by = "sample", 
                     all.x = TRUE)

# Add taxlevel to 'counts' by rankID from 'taxonomy'
counts_tidy_merged <- merge(counts_tidy_merged, taxonomy[, c("taxlevel", "rankID")],
                     by = "rankID",
                     all.x = TRUE)

# Filter rows where community is in Senegal and taxonomy level is 6
counts_selected <- counts_tidy_merged %>%
  filter(community %in% communities$original) %>%
  filter(taxlevel == 6)
```


## Question 6:
Calculate fractions of counts for each sample per genus
```{r question 6}
# Calculate total counts per sample and the fraction of counts in each sample per genus
counts_selected <- counts_selected %>% 
  group_by(sample) %>% 
  mutate(fraction = count / sum(count), total_count = sum(count)) %>% 
  ungroup()
```



## Question 7:
Calculate the total LAB fraction for each sample
```{r question 7, fig.cap="Figure 1. Frequency of the total fractions of LAB"}
#Merge the taxon column to data
counts_selected <- merge(counts_selected, 
                         taxonomy[, c("taxon", "rankID")],
                         by = 'rankID', 
                         all.x = TRUE)
  
counts_updated <- counts_selected %>%
  group_by(sample) %>% 
  filter(taxon=='Lactococcus'| taxon=='Lactobacillus'| taxon=='Streptococcus') %>% 
  mutate(LAB_total_fraction = sum(count) / total_count) %>% 
  ungroup()

#Histogram of the total fraction of these LAB
counts_updated %>%
  group_by(sample) %>%
  ggplot(mapping = aes(x = LAB_total_fraction)) +
  geom_histogram() +
  labs(x = "total fraction of LAB", y = "frequency", title = "Frequency of the total fraction of LAB")
```


## Question 8:
Summarise the number of samples per dominant LAB
```{r question 8}
# Make an object in which the dominant LAB genus is listed per sample
# If the total LAB is less than 90% then set the value for dominant LAB genus to “(none)”
counts_updated <- counts_updated %>%
  group_by(sample) %>% 
  mutate(dominant_LAB_genus = ifelse(LAB_total_fraction > 0.9, taxon[which.max(fraction)], "none")) %>% 
  ungroup()

# Print a table that summarizes the total number of samples per dominant LAB.
summary_tibble <- counts_updated %>% 
  group_by(sample, dominant_LAB_genus) %>% 
  summarise(count = n()) %>% 
  group_by(dominant_LAB_genus) %>% 
  summarise(count = n()) %>% 
  ungroup()

# Rename columns
summary_tibble2 <- summary_tibble
summary_tibble2 <- rename(summary_tibble2, "Dominant LAB genus" = dominant_LAB_genus)
summary_tibble2 <- rename(summary_tibble2, "#Samples" = count)

summary_tibble2 %>%
  knitr::kable(format="html", caption="Table 1. Number of samples in each dominant LAB genus.") %>% 
  kableExtra::kable_styling()
```




## Question 9:
Summarise the fraction of contaminated samples by the dominant LABs
```{r question 9}
# Get genera file
#current_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
#data_file_path <- paste(current_dir, '/data', sep = "")
#setwd(data_file_path)

genera <- read.csv('colon_genera.tab',
                   header = TRUE,
                   sep = '',
                   comment.char = '#')

genuses <- genera$genus

# Compute whether a sample is heavily contaminated (i.e. total fraction of colon > 0.1)
counts_contaminated <- counts_selected %>% 
  group_by(sample) %>% 
  filter(taxon %in% genuses) %>% 
  mutate(fraction_contaminated = sum(count) / total_count,
         contaminated = ifelse(fraction_contaminated > 0.01, TRUE, FALSE)) %>% 
  ungroup()

# Add the contaminated column to 'counts'
counts_contaminated_updated <- merge(counts_updated,
                                     counts_contaminated[, c("sample", "fraction_contaminated", "contaminated")],
                                     all.x = TRUE)

#Filter out duplicate sample rows
counts_contaminated_updated2 <- counts_contaminated_updated[!duplicated(counts_contaminated_updated$sample), ]

# Calculate the amount of contaminated samples
contaminated_summary <- counts_contaminated_updated2 %>% 
  group_by(dominant_LAB_genus, contaminated) %>%
  filter(contaminated == TRUE) %>% 
  summarise(count = n()) %>% 
  ungroup()

# Add the number and fraction of contaminated samples to each group
summary_tibble$number_contaminated <- contaminated_summary$count
summary_tibble$fraction_contaminated <- summary_tibble$number_contaminated / summary_tibble$count

# Rename columns
summary_tibble <- rename(summary_tibble, "Dominant LAB genus" = dominant_LAB_genus)
summary_tibble <- rename(summary_tibble, "#Samples" = count)
summary_tibble <- rename(summary_tibble, "#contaminated" = number_contaminated)
summary_tibble <- rename(summary_tibble, "Fraction contaminated" = fraction_contaminated)

#print(summary_tibble)
summary_tibble %>%
  knitr::kable(format="html", caption="Table 2. Number of samples and contaminated samples and their fraction of all samples in each dominant LAB genus.") %>% 
  kableExtra::kable_styling()

```


## Question 10:
```{r question 10, fig.cap="Figure 2. Total fraction of contaminating genera of samples as a function for the Shannon-Wiener diversity index"}
# graph displaying the total fraction of contaminating genera of samples as a function fo the Shannon-Wiener diversity index

counts_per_sample <- counts_contaminated_updated2 %>% 
  group_by(sample) %>% 
  mutate(shannon_wiener = -sum(fraction * log(fraction))) %>%
  ungroup()

counts_per_sample %>% 
  ggplot(mapping = aes(x = shannon_wiener,
                       y = fraction_contaminated,
                       color = dominant_LAB_genus)) +
  geom_point() +
  labs(x = "Shannon Wiener index", y = "Fraction of contaminated samples", title = "Total fraction of contaminating genera of samples as a function for the Shannon-Wiener diversity index")

```
#### comments:
- The above plot seems to have very low Shannon Wiener values so I've most likely made a mistake at some point. I'm aware of this and I would have fixed the issue if I wouldn't have ran out of time.



## Question 11
```{r question 11}
# t-test to investigate whether the total fraction of contamination differs significantly between samples that do and don’t have a dominant LAB
samples_dominant_LAB <- counts_per_sample[counts_per_sample$dominant_LAB_genus != "none", ]
samples_none <- counts_per_sample[counts_per_sample$dominant_LAB_genus == "none", ]

t_result <- t.test(samples_dominant_LAB$fraction_contaminated, samples_none$fraction_contaminated)

print(t_result)
```


#### Conclusion:  

t-test gives a low p-value, indicating that the null hypothesis should be rejected. Thus, the groups seem to differ in terms of the fraction of contaminated samples. In other words, it seems that there is a correlation between not having a dominant LAB genus and having more contaminated samples.


## Question 12
```{r}
#table that lists the dominant LAB categories per village (community)
village_table <- counts_per_sample %>% 
  group_by(community, dominant_LAB_genus) %>% 
  summarise(count = n())

village_table_new <- spread(data = village_table, dominant_LAB_genus, count)

village_table_new %>%
  knitr::kable(format="html", caption="Table 3. Dominant LAB categories per village") %>% 
  kableExtra::kable_styling()

```

  
  
#### comments:

- By looking at the data, it seems that 'Streptococcus' is the dominant LAB category for each village. 'Keur Mbaye Peul' has, however, equal amount of 'nones' as Streptococcuses. Also, 'None' seems to be the smallest group for all the other villages.

- It would be particularly interesting to further investigate the amount of each LAB category instead of just summarizing by results the most dominant LAB category as the differences might be rather small between the categories