---
title: "Statistics with R - Assignment1"
author: "Ville Lehtonen"
output: html_document
date: "`r strftime(Sys.Date(), '%B %d %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(knitr)
library(readxl)
library(tidyverse)
#current_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
#data_file_path <- paste(current_dir, '/data', sep = "")
#knitr::opts_knit$set(root.dir = data_file_path)

```

## Question 1:
#### What was good in the data
 - no cell formatting was used (i.e. colors)  
 
 - comments were put in separate columns  
 
 - variable names were clear and camelCase was used mostly (although in some columns this was not the case)  
 
 - Data was split into different tables and samples and counts tables were connected by the sample value  
 
#### What could be improved
 - The counts data could have been in a 'long' format i.e. there should have been a column called 'sample' and 'count', instead of having multiple samples as separate columns  
 
 - The 'bad data' (with i.e. missing values) could have been labeled in a new column that would have had just TRUE / FALSE values indicating whether this is good or bad data. This would ease the filtering of data  

 - in 'samples' table, the time columns could have been split into hours and minutes columns so that they would only include numerical values  
 
- in 'samples' table, there was a column called 'duplicate' that contained french (Oui / Non). Instead it would be simpler if it would contain only a logical TRUE / FALSE value  

- Metadata was missing  


## Question 2:
Create objects for counts, samples, and communities data
```{r question 2}
counts <- read_xlsx('lait_caille_data.xlsx',
                    sheet = 'counts',
                    col_names = TRUE)

samples <- read_xlsx('lait_caille_data.xlsx',
                    sheet = 'samples',
                    col_names = TRUE)

communities <- read_xlsx('lait_caille_data.xlsx',
                    sheet = 'communities',
                    col_names = TRUE)

```


## Question 3:
Split counts data and tidy it
```{r question 3}
# Split counts data into taxonomy and counts objects (both contain rankID)
taxonomy <- select(counts,taxlevel, rankID, taxon)
counts <- select(counts, -taxlevel, -taxon)

# Transform counts object into a tidy format with 3 columns (rankID, sample, count)
counts_tidy <- gather(data = counts,
                      key = sample,
                      value = count,
                      -rankID)
```


## Question 4:
Perform checks
```{r question 4}
# Gather all unique sample IDs from 'counts' & 'samples'
counts_IDs <- unique(unlist(counts_tidy$sample, " "))
samples_IDs <- samples$sample

# Check that both IDs vectors contain the same objects
print(c("Check that all IDs of counts are in samples:", all(counts_IDs %in% samples_IDs)))
print(c("Check that all IDs of samples are in counts:", all(samples_IDs %in% counts_IDs)))

# Gather unique community names from 'communities' and 'samples'
original_names <- communities$original
samples_communities <- unique(unlist(samples$community, " "))

# Check if all original names are included in communities in 'samples' and vice versa 
print(c("Check that all original names are in the communities of samples:", all(original_names %in% samples_communities)))
print(c("Check that all communities of samples are in the original names:", all(samples_communities %in% original_names))) #-> FALSE

# Gather the missing communities (= surplus or have missing values) into a list
missing_communities <- samples_communities[!samples_communities %in% original_names]
```


## Question 5:
Filter rows where community is in senegal and taxonomy level is 6
```{r question 5}
# Add community to 'counts' by sample ID from 'samples'
counts_tidy_merged <- merge(counts_tidy, samples[, c("sample", "community")], 
                     by = "sample", 
                     all.x = TRUE)

# Add taxlevel to 'counts' by rankID from 'taxonomy'
counts_tidy_merged <- merge(counts_tidy_merged, taxonomy[, c("taxlevel", "rankID")],
                     by = "rankID",
                     all.x = TRUE)

# Filter rows where community is in Senegal and taxonomy level is 6
counts_selected <- counts_tidy_merged %>%
  filter(community %in% communities$original) %>%
  filter(taxlevel == 6)
```


## Question 6:
Calculate fractions of counts for each sample per genus
```{r question 6}
# Calculate total counts per sample and the fraction of counts in each sample per genus
counts_selected <- counts_selected %>% 
  group_by(sample) %>% 
  mutate(fraction = count / sum(count), total_count = sum(count)) %>% 
  ungroup()
  
```

## Question 7:
Calculate the total LAB fraction for each sample
```{r question 7}
#Merge the taxon column to data
counts_selected <- merge(counts_selected, 
                         taxonomy[, c("taxon", "rankID")],
                         by = 'rankID', 
                         all.x = TRUE)
  
counts_updated <- counts_selected %>%
  group_by(sample) %>% 
  filter(taxon=='Lactococcus'| taxon=='Lactobacillus'| taxon=='Streptococcus') %>% 
  mutate(LAB_total_fraction = sum(count) / total_count) %>% 
  ungroup()

#Histogram of the total fraction of these LAB
#hist(counts_updated$LAB_total_fraction)

# counts_updated %>%
#   group_by(sample) %>% 
#   ggplot(mapping = aes(x = sample,
#                        y = LAB_total_fraction)) +
#   geom_col()
```


## Question 8:
Summarise the number of samples per dominant LAB
```{r question 8}
# Make an object in which the dominant LAB genus is listed per sample
# If the total LAB is less than 90% then set the value for dominant LAB genus to “(none)”
counts_updated <- counts_updated %>%
  group_by(sample) %>% 
  mutate(dominant_LAB_genus = ifelse(LAB_total_fraction > 0.9, taxon[which.max(fraction)], "none")) %>% 
  ungroup()

# Print a table that summarizes the total number of samples per dominant LAB.
summary_tibble <- counts_updated %>% 
  group_by(sample, dominant_LAB_genus) %>% 
  summarise(count = n()) %>% 
  group_by(dominant_LAB_genus) %>% 
  summarise(count = n()) %>% 
  ungroup()

print(summary_tibble)
```




## Question 9:
Summarise the fraction of contaminated samples by the dominant LABs
```{r question 9}
# Get genera file
#current_dir <- dirname(rstudioapi::getActiveDocumentContext()$path)
#data_file_path <- paste(current_dir, '/data', sep = "")
#setwd(data_file_path)

genera <- read.csv('colon_genera.tab',
                   header = TRUE,
                   sep = '',
                   comment.char = '#')

genuses <- genera$genus

# Compute whether a sample is heavily contaminated (i.e. total fraction of colon > 0.1)
counts_contaminated <- counts_selected %>% 
  group_by(sample) %>% 
  filter(taxon %in% genuses) %>% 
  mutate(fraction_contaminated = sum(count) / total_count,
         contaminated = ifelse(fraction_contaminated > 0.01, TRUE, FALSE)) %>% 
  ungroup()

# Add the contaminated column to 'counts'
counts_contaminated_updated <- merge(counts_updated,
                                     counts_contaminated[, c("sample", "fraction_contaminated", "contaminated")],
                                     all.x = TRUE)

#Filter out duplicate sample rows
counts_contaminated_updated <- counts_contaminated_updated[!duplicated(counts_contaminated_updated$sample), ]

# Calculate the amount of contaminated samples
contaminated_summary <- counts_contaminated_updated %>% 
  group_by(dominant_LAB_genus, contaminated) %>%
  filter(contaminated == TRUE) %>% 
  summarise(count = n()) %>% 
  ungroup()

# Add the number and fraction of contaminated samples to each group
summary_tibble$number_contaminated <- contaminated_summary$count
summary_tibble$fraction_contaminated <- summary_tibble$number_contaminated / summary_tibble$count

print(summary_tibble)
```


## Question 10:
```{r question 10}
# graph displaying the total fraction of contaminating genera of samples as a function fo the Shannon-Wiener diversity index


# ADD SHANNON INDEX!
counts_contaminated_updated %>%
  group_by(sample) %>% 
  ggplot(mapping = aes(x = sample,
                       y = fraction_contaminated,
                       fill = dominant_LAB_genus)) +
  geom_col()
```

## Question 11
```{r question 11}
# t-test to investigate whether the total fraction of contamination differs significantly between samples that do and don’t have a dominant LAB

```


























